---
title: "Demonstration of netCDF functionality in rSW2analysis"
author: "Caitlin M. Andrews"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Demonstration of netCDF functionality in rSW2analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Overview

This vignette is an overview of how to use the functions in the rSW2analysis package to
convert arrays of rSOILWAT2 and SOILWAT2 output data into netCDFs that adhere to
[CF conventions](http://cfconventions.org/Data/cf-conventions/cf-conventions-1.8/cf-conventions.html#_orthogonal_multidimensional_array_representation).

Additionally, this document describes the Dryland Ecology Lab's standards and
conventions in regards to naming, metadata, and netCDF structure.

# Input Data Format and Structure

## Variable Data

The netCDF functions in this package expect an array, in which each row represents
data for a site. Columns represent one of three things:

* multiple different variables for one site
* one variable with multiple measurements (time series)
* one variable with multiple depth measurements

```{r, create1d, results = 'hide'}
   # load rSW2analysis pacckage
    library(rSW2analysis)

   # dummy data for 10 sites, 10 data points (could be 10 vars, 10 times, or 10 depths)
   someData <- rnorm(100, 7, 30)
   data1d <- array(someData, c(10, 10))
   str(data1d)
   is.array(data1d)
```

Additionally, it is possible to have data at multiple times *and* multiple depths.
In this case, time should be organized by column, and multiple depths should be organized
sequentially in third dimension, forming a 3d array.

```{r, create3d, results = 'hide'}
  # dummy data for 10 sites, 10 times, 3 depths.
  data3d <- array(0, c(10, 10, 3))
  data3d[,,1] <- someData
  data3d[,,2] <- someData
  data3d[,,3] <- someData
```

## Location Data

Locations all need to be on the same coordinates reference system (CRS). Data can
be either be organized on a regularly spaced grid (i.e. wall to wall simulations for
a given area) or not (i.e. project data representing plot points in a park).

If the former, the coordinates need to be in order. If the later,

```{r, createlocs, results = 'hide'}
  # points on a regular spaced grid in the WGS84 CRS
  locationsGrid <- data.frame(X_WGS84 = c(rep(-124.5938, 5), rep(-124.5312, 5)),
                        Y_WGS84 = rep(c(47.90625, 47.96875, 48.03125, 48.09375,
                                     48.15625), 2))

  # irregularly spaced points in the WGS84 CRS
  #locationsRandom
```

## Dimension Data

NetCDFs are organized into dimensions. Dimensions have two key components: Bounds
and values. The dimensions also have other information which is degined in the Attibute Data
(see below).

### Latitude (Y) and Longitude (X) Dimensions

All NetCDFs will have at least two dimensions, latitude and longitude.

The bounds and values of the latitude and longitude are determined from the location
data within the *create_empty_netCDF_file* function.

### Time (T) Dimension

The user must define the bounds and values for the time dimension, as inputs to the
*create_empty_netCDF_file*.

The bounds of the time dimension are the start and end of each interval of the time dimension.

The values refer to the midpoint of each interval of the time dimension.

The bounds and values need to be in the same units.

A common unit for netCDF is days since 1900. A full list of accepted time units can
be found [here](https://ncics.org/portfolio/other-resources/udunits2/).

```{r, timeD}

    ## time vals
    time_midpoints <- c(43920, 44285, 44650, 45015, 45381, 45746, 46111,
                          46476, 46842, 47207)# mid point of year n days since 1900-01-01

    ## time bounds
    annual_time_bounds <- c(43737, 44102, 44102, 44467, 44467, 44832, 44833, 45197,
                   45197, 45563, 45563, 45928, 45928, 46293, 46293, 46658,
                  46658, 47024, 47024, 47389) # beginning and end of year in days since 1900-01-01
```

### Depth (Z) Dimension

The user must define the bounds and values for the time dimension.

```{r, depthD}

    ## depth vals
    depths_max <- c(5, 10, 15) # maximum depth of each layer of depth

    ## depth bounds
    depth_bounds <- c(0, 5, 5, 10, 10, 15) # min and max of each nlayer of depth depth
```

## Attribute Data

Attribute data pertains to the metadata contained in the netCDFs. NetCDFs are
self-documenting and require metadata attributes, providing information
necessary for interpreting the data.

Different attribute types include variable attributes, time attributes, depth
attributes, and global attributes. The format of attributes is a named list. Each attribute list
has a bare minimum of required fields. Endless other fields are optional and can
be added by the user as they see fit.

### Variable Attributes

A named list describing the variable data. The *name* and *units* fields are required.

```{r, VarAtt, results = 'hide'}
# single variable
var_attributes <- list(
  name = 'WatYrWDD',
  long_name = 'Water Year Wet Degree Days',
  units = 'number of days'
)

# multiple variables
var_attributes2 <- list(
    name = paste0('VarDD_', 1:10),
    units = rep('number of days', 10)
)

```

### Time Attributes

A named list describing the time dimension. The *name*, *units*, *calender*, and *vals*
fields are required here.

A common unit for netCDF is days since 1900. A full list of accepted time units can
be found [here](https://ncics.org/portfolio/other-resources/udunits2/).

```{r, TimeAtt, results = 'hide'}
time_attributes <- list(
  name = 'time',
  units = 'days since 1900-01-01',
  calendar = 'standard',
  unlim = TRUE,
  vals = time_midpoints # mid point of year
  )
```

### Vertical (depth / Z) Attributes

A named list describing the vertical (depth / Z) dimension. The *name*, *units*,
and *vals* fields are required.

```{r, VertAtt, results = 'hide'}
vertical_attributes <- list(
  name = 'depth',
  units = 'cm',
  positive = 'down', # depth is down, height is up
  vals = depths_max
)
```

### Global Attributes

A named list describing the global attributes of the netCDF. This is where you describe
things like how the data was made, where the data was made, who the data was made by,
what climate models were used to create the data, etc.

The *title*, *institution*, *description*, *source_id*, *realm*, *product*, and
*contact* fields are required.

```{r, GlobalAtt, results = 'hide'}
global_attributes <- list(
    title = "Project Name Annual Wet Degree Days",
    institution = 'Southwest Biological Science Center, U.S. Geological Survey',
    description = 'methods on how this data was made',
    source = paste(
      "SOILWAT2 (v4.2.0);",
      "rSOILWAT2 (v2.3.2);",
      "rSFSW2 (v3.1.2)"
    ),
    source_id = "SOILWAT2",
    realm = "land",
    parent_mip_era = "CMIP5",
    parent_experiment_id = "RCP45",
    parent_source = "CanESM2",
    parent_variant_label = "r1i1p1f1",
    product = "model-output",
    projection = 'Geographic',
    grid = 'WGS84',
    grid_label = "gn",
    nominal_resolution = "10 km", # \code{\link{populate_netcdf_from_array}}
    further_info_url = "https://github.com/DrylandEcology/",
    contact = "you@email.com"
    )
```

#### Nominal resolution

If the data is calculated and reported on a grid, CF convention calls for the
field *nominal resolution* to be contained in the global attributes. The function
*calculate_nominal_resolution* will calculate the nominal resolution of your gridded
data for you. However, you need to provide a raster of the grid, the lat/long locations, and
the area that each cell in the grid represents.

```{r, Nominal, results = 'hide'}
r <- raster::raster(
  xmn = -120, xmx = -90,
  ymn = 30, ymx = 50,
  crs ="+init=epsg:4326",
  resolution = c(0.5, 0.5)
)
r[] <- seq_len(prod(dim(r)))
xy <- raster::sampleRandom(r, size = 50, sp = TRUE)
grid_cell_area <- calculate_cell_area(sites = xy, grid = r)

calculate_nominal_resolution(
  grid = r,
  sites = xy,
  cell_areas_km2 = grid_cell_area[, "km2"]
)
```

# Output Files Structure Naming Conventions

File structure should be organized by: variableName / allFilesPerVariable. Each file
will contain variable output for one time period and/or climate model.

The name for each file should be in the format as follows:
**var_LorA_timestep_Project_SOILWAT2_GCM_RCP_variantLabel_gn_startT_endT.nc**

where:

* var – Shorthand name for variable that is contained in files (i.e. ppt for precipitation, WDD for wet degree days)
* LorA – Land (L) or atmospheric (A) data
* timestep – The time-step of the data contained (i.e. the units of the time dimension (i.e. year, month)).
* Project – Project Name (i.e. SEUG or RangeResil, etc.)
* SOILWAT2 – Keep as is. Denotes that SOILWAT2 was used to generate these results
* GCM – Name of the GCM climate data used to drive the SOILWAT2 model (i.e. IPSL5, CCSM2-1)
* RCP – Name of which GCM used to drive the GCM (RCP4.5, RCP8.5)
* variantLabel –Variant of the GCM used (i.e. r1i1p1f1)
* gn
* startT – lower time bound or start of the time period contained. Units should be the date (YearMonthDay, XXXXYYZZ).
* endT – upper time bound or end of the time period contained. Units should be the date (YearMonthDay, XXXXYYZZ).

# Create netCDF Function and Examples

The functions create_empty_netCDF_file and populate_netcdf_from_array will create and populate
a netCDF from an array. The user defines the dimensions, attribtues, and locations in the functions'
arguments.

The two functions work in conjuction with one another. First, an empty netcdf is created.
Then, it is filled with data.

## Example 1 - Create a 3 dimension netCDF (lat, long, time) from gridded data
```{r, 3dGridExample, results = 'hide'}
    ## locations
    crs <- '+init=epsg:4326' # the CRS

    outFileName <- 'dummynetcdf.nc'

    # create
    create_empty_netCDF_file(
      data = data1d,
      has_T_timeAxis = TRUE,
      has_Z_verticalAxis = FALSE,
      time_bounds = annual_time_bounds,
      vert_bounds = NULL,
      var_attributes = var_attributes,
      time_attributes  = time_attributes,
      vertical_attributes = NULL,
      global_attributes = global_attributes,
      isGridded = TRUE,
      locations = locationsGrid,
      crs = crs,
      file = outFileName,
      force_v4 = TRUE,
      overwrite = TRUE
    )

    # populate
    populate_netcdf_from_array(
      file = outFileName,
      data = data1d,
      var_names <- var_attributes$name,
      has_T_timeAxis = TRUE,
      has_Z_verticalAxis = FALSE,
      isGridded = TRUE,
      locations = locationsGrid,
      crs = crs,
      force_v4 = TRUE
    )

```
## Example 2 - Create a 2 dimension netCDF (lat, long) from gridded data

```{r, 2dExample, results = 'hide'}

    # create
    create_empty_netCDF_file(
      data = data1d,
      has_T_timeAxis = FALSE,
      has_Z_verticalAxis = FALSE,
      time_bounds = NULL,
      vert_bounds = NULL,
      var_attributes = var_attributes2,
      time_attributes  = NULL,
      vertical_attributes = NULL,
      global_attributes = global_attributes,
      isGridded = TRUE,
      locations = locationsGrid,
      crs = crs,
      file = outFileName,
      force_v4 = TRUE,
      overwrite = TRUE
    )

    # populate
    populate_netcdf_from_array(
      file = outFileName,
      data = data1d,
      var_names <- var_attributes2$name,
      has_T_timeAxis = FALSE,
      has_Z_verticalAxis = FALSE,
      isGridded = TRUE,
      locations = locationsGrid,
      crs = crs,
      force_v4 = TRUE
    )
```

## Example 3 - Create a 3 dimension netCDF (lat, long, time) from non-gridded data

```{r, 3dNoGridExample, results = 'hide'}
# to do
```

## Example 4 - Create a 4 dimension netCDF (lat, long, time, depth) from gridded data

```{r, 4dGridExample, results = 'hide'}

 create_empty_netCDF_file(
      data = data3d,
      has_T_timeAxis = TRUE,
      has_Z_verticalAxis = TRUE,
      time_bounds = annual_time_bounds,
      vert_bounds = depth_bounds,
      var_attributes = var_attributes,
      time_attributes  = time_attributes,
      vertical_attributes = vertical_attributes,
      global_attributes = global_attributes,
      isGridded = TRUE,
      locations = locationsGrid,
      crs = crs,
      file = outFileName,
      force_v4 = TRUE,
      overwrite = TRUE
    )

     # populate
    populate_netcdf_from_array(
      file = outFileName,
      data = data3d,
      var_names <- var_attributes$name,
      has_T_timeAxis = TRUE,
      has_Z_verticalAxis = TRUE,
      isGridded = TRUE,
      locations = locationsGrid,
      crs = crs,
      force_v4 = TRUE
    )
```
